---
title: "Benthos Demonstration"
output: html_notebook
---

Benthos utilizes several highly-technical indices/functions which can feel extremely cumbersome to non-professionals like us, so I will demonstrate some of the more simple functions that it offers.

To begin, please install.packages("benthos") and run the code below:
```{r}
install.packages('benthos')
library(benthos)
#the following are for data manipulation/visualization:
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
```

We will illustrate the benthos-package by means of the Oosterschelde marine benthos data set. This data set ships with the benthos-package and can be loaded by typing:
```{r}
data("oosterschelde")
#type ?oosterschelde if you want to see documentation on data
```
 
 
 
 We can then **preprocess** the data for analysis simplification; for example, say we want to view the samples (stored in rows) taken in August and September:
```{r}
oosterschelde <- oosterschelde %>%
    filter(months(DATE) %in% c("August", "September"))
```

..Or we can create a new column by combining data from the samples into 'YEAR':
```{r}
oosterschelde <- oosterschelde %>%
    mutate(YEAR = DATE %>% format("%Y"))
oosterschelde
```


Several biodiversity measures can be used with the pre-processed data, such as...


The **total abundance** is the total number of individuals in a sampling unit:
```{r}
oosterschelde %>% total_abundance(count=COUNT)
```

The **abundance** is the total number of individuals per taxon in a sampling unit (Maybe don't run this one):
```{r}
oosterschelde %>% abundance(taxon=TAXON, count=COUNT) %>% as.matrix
```


**Species richness, S** is the number of different species in a (pooled) sample:
```{r}
oosterschelde %>% species_richness(taxon=TAXON, count=COUNT)
```


**Simpson’s Measure of Concentration** gives the probability that two individuals selected at random from a sample will belong to the same species. "For an infinite sample Simpson’s Index is given by:
λ=∑i=1Sπ2i
where πi the proportion of the individuals in species i. For a finite sample, Simpson’s index is:
L=∑i=1Sni(ni−1)N(N−1)
where ni the number of individuals in species i and N the total number of individuals."

The finite sample case can be computed by:
```{r}
oosterschelde %>% simpson(taxon=TAXON, count=COUNT)
```
