---
title: "Tidy Data"
---
```{r}
installed.packages('tidyr')
install.packages('tidyr')
```


```{r setup}
library(tidyverse)

# Toy data
cases <- tribble(
  ~Country, ~"2011", ~"2012", ~"2013",
      "FR",    7000,    6900,    7000,
      "DE",    5800,    6000,    6200,
      "US",   15000,   14000,   13000
)

pollution <- tribble(
       ~city, ~size, ~amount,
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",     121,
   "Beijing", "small",     121 #'untidy' when data is in this format
)

names(who) <- stringr::str_replace(names(who), 
                                   "newrel", 
                                   "new_rel") # R reads columns as variable names
```

```{r}
table1 #country, year, cases, pop
table2 #country, year, type, count
table3 #country, year, rate
table4a #country, 1999, 2000
table4b #country, 1999, 2000 (large numbers under years)
table5
```

# tidyr

# Data is 'tidy' iff: variables are columns, obs are rows, each cell has 1 value

## Your Turn 1

On a sheet of paper, draw how the cases data set would look if it had the same values grouped into three columns: **country**, **year**, **n**

# key = former column name
# value = former cells
# new names are in quotes


Try the following codes - what do they do? (annotate them with a comment)

```{r}
cases %>% gather(key="year",value="n", 2:4)
#can also be written as cases %>% gather("year",n", 2:4) #works when cloumns are next to e/other
cases %>% gather(key="year",value="n", "2011", "2012", "2013") #can get tediousto write every year 
cases %>% gather(key="year",value="n", starts_with("201")) #good for data spread out among columns
cases %>% gather(key="year",value="n", -Country) #4 different ways of doing the same thing
#all charts are the same!!
```




tidyr is transitioning `gather()` to `pivot_longer()`.  Here are equivalent codes:
```{r}
cases %>% gather(key="year",value="n", 2:4) #gather is the new update
cases %>% pivot_longer(2:4, names_to = "year", values_to = "n") 
```


## Your Turn 2
Rewrite the following codes as `pivot_longer()` statements
```{r}
cases %>% gather(key="year",value="n", 2:4)
cases %>% pivot_longer(2:4, names_to = "year", values_to = "n")
cases %>% gather(key="year",value="n", "2011", "2012", "2013")
cases %>% pivot_longer(names_to = "year", values_to = "n", c("2011", "2012", "2013")
cases %>% gather(key="year",value="n", starts_with("201"))
cases %>% pivot_longer(names_to ="year", values_to ="n", starts_with("201"))
cases %>% gather(key="year",value="n", -Country)
cases %>% pivot_longer(names_to="year",values_to ="n", -Country)
#change gather --> "pivot_longer", key --> names_to, value --> values_to
```





## Your Turn 3

Use `gather()` or `pivot_longer()` to reorganize `table4a` into three columns: **country**, **year**, and **cases**.

```{r}
#ctrl + shift + M inserts %>% 
table4a %>% pivot_longer(col=2:3, names_to = "year", values_to = "n", names_ptypes = list(year=integer())) # n stands for case
# year is stored as chr until r is told to do otherwise
# for gather command use "convert"
#once data has bee named it doesn't need quotes
```



## Your Turn 4

On a sheet of paper, draw how this data set would look if it had the same values grouped into three columns: **city**, **large**, **small**

## Your Turn 5


Use `spread()` or `pivot_wider()` to reorganize `table2` into four columns: **country**, **year**, **cases**, and **population**.

```{r}
table2 %>% 
  spread(key = type, value=count) #includes as many cloumns as are in type & values from count
table2 %>% 
  pivot_wider(names_from = type, values_from = count)
```

# who 
#variable names have 3 types of data included in them: type, sex, & age
#iso names are redundant info so we don't need them


## Your Turn 6

Gather the 5th through 60th columns of `who` into a pair of key:value columns named `codes` and `n`.

Then select just the `county`, `year`, `codes` and `n` variables.

```{r}
who %>%
  gather(key = codes, value = n, 5:60) %>%
  select(-iso2, -iso3) #dplyr fx to remove this info since it's redundant 
```


Bonus: Try to do the same code with `pivot_longer()`
```{r}
who %>%
  pivot_longer(names_to = _____, values_to =____, cols=5:60) %>%
  select(-iso2, -iso3)

```


## Your Turn 7

Separate the `sexage` column into `sex` and `age` columns.

*(Hint: Be sure to remove each `_________` before running the code)*

```{r}
who %>%
  gather("codes", "n", 5:60) %>%
  select(-iso2, -iso3) %>%
  separate(codes, c("new", "type", "sexage"), sep = "_") %>%
  select(-new) %>% 
  _______________________________
```


***

# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.

